---
output: 
  html_fragment
bibliography: ptpt.bib
---

```{r render,include = FALSE,echo = FALSE}
# Need to fix toc formatting - probably in css or header/footer files
```

<span style="color: red;">**NOTE**</span>

This documentation is still under review. Be aware that the content may change.
Check the github commit history for this project's [repo](github.com/apeterson91/ptpt)
to see when changes are made.

# Preface

This document contains information describing the context, motivation and 
technical implementation of the Pittsburgh Transit Propensity Tool. As we 
aspire to have this document be accessible both to the lay and technical 
public alike, each section is tabbed into two sections: "Big Picture" and 
"Technical". The former is written for a lay person interested in understanding
this project from a "Big Picture" perspective and consequently efforts are made 
to remove as much technical jargon, citations or mathematical notation to ensure
reader understanding. In contrast, the Technical tab is written for a research
audience looking to understand, comment on and possibly build off of this work.

# Introduction {.tabset}

## Big Picture

Cars are expensive. Whether considering the financial cost of buying, 
insuring, fueling and maintaining a car, the cost of human life and injury 
caused annually by vehicle accidents, their contribution to environmental 
problems like global warming and air pollution, or more aesthetic concerns 
regarding noise and city views, the current state of car-dependency that is 
common in much of the U.S. presents a serious problem for the economic, 
physical and mental well-being of its current and future residents.

Charting a path from the current state of infrastructure in U.S. cities to a 
place of independence from cars is no simple feat, however. The car has been 
the dominant mode of transit in the U.S. for decades and zoning 
and land use patterns have reinforced the place of the car as a necessity
within the average American's daily life. Decoupling the car from everyday life
will then require some measure of understanding where and why individuals *must*
use cars so that policy makers can more precisely target funding for
transit, pedestrian and bike infrastructure and incentives.

### The Pittsburgh Transit Propensity Tool (PTPT)

### Site and Methods Outline


## Technical

Cars are expensive. Whether considering the financial cost of buying, 
insuring, fueling and maintaining a car, the cost of human life and injury 
caused annually by vehicle accidents, their contribution to environmental 
problems like global warming, or simple aesthetic concerns regarding noise, 
and looks, the current state of car-dependency that is common in much of the 
U.S. presents a serious problem for the economic, physical and mental well-being
of its current and future residents.

Charting a path from the current state of infrastructure in U.S. cities to a 
place of independence from cars is no simple feat, however. The car has been 
the dominant mode of transit in the U.S. for decades (cite census data) and zoning 
and land use patterns have reinforced the place of the car as a necessity
within the average American's daily life [@guillen2020motorisation] . 
Take a different direction from here?

With this principle in mind, we propose to develop and adapt the the Propensity
to Cycle Tool (PCT) [@lovelace2017propensity] which offers an intuitive way to 
understand where and to what extent residents of the UK 
utilize cycling as a mode of transit for their daily work commute. Our project 
aims to adapt and extend the work of the PCT to the city of Pittsburgh in 
order to map how and where individuals move around the city.

# Data {.tabset}

## Big Picture

A number of data sources are used in the construction of this project. 
We provide a list below in Table 1.

```{r datasources,message = FALSE, warning = FALSE, echo = FALSE}
library(dplyr)
library(gt)
foo <- tibble(Source = c("WPRDC","WPRDC","WPRDC","OSM"),
              Name = c("Street Level Data",
                       "Parcel Data",
                       "Neighborhood Locations",
                       "Grocery Store Locations"))
foo %>% gt()
```

## Technical

# Methods 

## Stage 1 {.tabset}

### Big Picture

### Technical

We generate 100 subjects from each of the 88 Pittsburgh neighborhoods
resulting in $n=880$ subjects. For each subject we generate a trip to a 
local grocery store by randomly selecting one of the 3 closest grocery stores
near the residents' parcel 

## Stage 2 {.tabset}

### Big Picture

### Technical

# Snippets

Perhaps the greatest challenge, as with any abstract endeavor, is setting out 
a normative standard by which 
to orient one's efforts; what does a car-independent lifestyle look like? 
While Amsterdam, Copenhagen, New York City and other cities, offer several several 
points of reference in their use of bike infrastructure, high density zoning 
and public transit. Furthermore they show that a city need not offer a 
car-free solution to every trip that a resident might take, but rather 
focus on the *majority* of the trips that an individual takes --- those trips 
to the grocery store, work, urban green space and commercial or public spaces 
that form the bulk of trips for most.

# References
